---
title: "Scagnostic-based preprocessing for projection pursuit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scagnostic-based preprocessing for projection pursuit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
code-fold: true
warning: FALSE
message: FALSE
---

```{r setup, include=FALSE}
library(bilbyr)
library(cassowaryr)
library(dplyr)
library(ggplot2)
```

## Why preprocess high-dimensional data?

Finding interesting projections in high-dimensional data can be
computationally challenging. Many candidate projections may resemble
Gaussian noise, while informative structures may involve only a small
subset of variables.

`bilbyr` provides `scag_preprocess()` to calculate a selected
scagnostic for every pair of numeric variables and rank the pairs
in descending order of their scores.

These rankings provide an initial guide for projection-pursuit
optimizers, potentially helping them prioritize informative regions
of the search space.

## Exploring a real dataset

We begin with the wide-format Datasaurus dataset from `cassowaryr`.

```{r}
data("datasaurus_dozen_wide", package = "cassowaryr")

x <- datasaurus_dozen_wide[, c(
  "star_x", "star_y",
  "circle_x", "circle_y"
)]

prep <- scag_preprocess(
  x,
  scag = "stringy05",
  rescale = TRUE,
  binner = NULL,
  out.rm = FALSE
)

prep$pair_scores
```

The function evaluates every unique variable pair and places the
highest-scoring pair first. This allows us to identify potentially
interesting bivariate views without manually inspecting every
scatterplot.

## Can preprocessing recover a hidden pattern?

We now generate an eight-dimensional dataset containing six Gaussian
noise variables and a sine-wave relationship hidden in `V3` and `V4`.

```{r}
set.seed(1050)

n <- 500
p <- 8

X <- as.data.frame(matrix(rnorm(n * p), ncol = p))
names(X) <- paste0("V", 1:p)

t <- seq(-1, 1, length.out = n)

X$V3 <- t + rnorm(n, sd = 0.005)
X$V4 <- sin(2 * pi * t) + rnorm(n, sd = 0.005)

ggplot(X, aes(V3, V4)) +
  geom_point(size = 0.6) +
  theme(aspect.ratio = 1) +
  theme_bw()
```

The relationship is clearly visible when the correct variables are
plotted, but it is hidden among the other dimensions.

We apply preprocessing using rescaled Stringy05 index:

```{r}
prep_signal <- scag_preprocess(
  X,
  scag = "stringy05",
  rescale = TRUE,
  binner = NULL,
  out.rm = FALSE
)

head(prep_signal$pair_scores, 10)
```

We can examine the ranking of the true signal directly:

```{r}
signal_rank <- prep_signal$pair_scores |>
  filter(var1 == "V3", var2 == "V4")

signal_rank
```

For this simulation, the sine-wave pair receives rank
`r signal_rank$rank` out of `r choose(p, 2)` possible pairs.

A high ranking indicates that preprocessing successfully prioritizes
the hidden structure. However, this result depends on the pattern,
noise level, and scagnostic used; recovering every interesting
structure is not guaranteed.

## Why is noise rescaling important?

A useful projection pursuit index (PPI) should assign high scores to
interesting patterns and comparatively low scores to unstructured
noise.

This is particularly important in high-dimensional data, where
exploring many projections can produce numerous approximately
Gaussian bivariate views.

The original Stringy05 index can assign positive scores to Gaussian
noise. To account for this, we use a sample-size-dependent noise
cutoff estimated from the 95th percentile of Gaussian simulations:

$$
l(n) = 0.05 + \frac{3.86}{\sqrt{n}}.
$$

The rescaled score is

$$
S^* = \max\left(0,\frac{S-l(n)}{1-l(n)}\right).
$$

Scores below the estimated cutoff become zero, while scores above
it remain positive after rescaling.

### Comparing original and rescaled Stringy05

We generate eight independent Gaussian noise variables and evaluate
both versions on exactly the same data.

```{r}
set.seed(1050)

X_noise <- as.data.frame(matrix(rnorm(n * p), ncol = p))
names(X_noise) <- paste0("V", 1:p)

prep_raw <- scag_preprocess(
  X_noise,
  scag = "stringy05",
  rescale = FALSE,
  binner = NULL,
  out.rm = FALSE
)

prep_rescaled <- scag_preprocess(
  X_noise,
  scag = "stringy05",
  rescale = TRUE,
  binner = NULL,
  out.rm = FALSE
)
```

Join the results to compare the two scores for each pair:

```{r}
comparison <- prep_raw$pair_scores |>
  select(var1, var2, raw = score) |>
  inner_join(
    prep_rescaled$pair_scores |>
      select(var1, var2, rescaled = score),
    by = c("var1", "var2")
  )

comparison
```




Rescaling suppresses scores that fall below the estimated
Gaussian-noise threshold, reducing the prominence of these views.
It does not guarantee that every noise view receives a zero score:
the cutoff is calibrated for individual Gaussian views, and some
noise pairs may exceed it by chance.

Because all pairs have the same sample size, rescaling preserves
their score ordering except where values are tied at zero.

## Implications for projection pursuit

These experiments illustrate two complementary goals of preprocessing:
prioritizing structured variable pairs and suppressing the scores of
unstructured Gaussian views.

In projection pursuit, assigning low scores to noise-like views may
reduce their attractiveness to an optimizer, allowing more search
effort to be directed toward interesting structures. The highest-ranked
pairs could also provide informed starting projections.

However, pairwise preprocessing considers only coordinate-aligned
views. Interesting structures may exist in combinations of variables
that are not apparent in any original pair. Therefore, pair rankings
are intended to guide, rather than replace, the optimizer's search.

Whether preprocessing improves optimization speed and recovery will
be evaluated separately through repeated simulations.
